---
title: "Lead Contamination Testing: Is Water in Toronto Safe for Drinking?"
subtitle: "Water in Some Districts is More Contaminated Than Others"
author: 
  - Amy Jin
thanks: "Code and data are available at: https://github.com/aj3616/Lead-Testing-Program-in-Toronto"
date: today
date-format: long
abstract: "The City of Toronto offers a free, non-regulated Residential Lead Testing Program, where residents collect and submit water samples. Quality of drinking water is an important public health issue, since too much abosorbtion of lead might have detrimental effects especially for young children. This paper examines how time and district affects the lead concentration with respect to safe lead concentration guidelines. It was found that the lead concentration does not have noticeble seasonality or autocorrelation, but some areas in toronto tends to have higher lead concentrations than the rest."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(knitr)
library(gt)
library(lubridate)  # For handling date extraction
library(dplyr)

data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
lead = data$`Lead Amount (ppm)`
district = data$PartialPostalCode

blueishgrey <- rgb(96, 139, 182, maxColorValue = 255)

```

# Introduction

Lead contamination in drinking water poses significant public health
risks, particularly for vulnerable populations such as children and
pregnant women(@c-1).

This paper takes in raw data from the City of Toronto(Open Data
Toronto), where they implemented a non-regulated program called City's
Residential Lead Testing Program. More information on this program can
be explored at: www.toronto.ca/leadpipes. The results respond to the
concerns about lead exposure in the city of Toronto.

While significant research has documented the public health
impacts of lead exposure, particularly focusing on its toxic effects on
vulnerable populations like children and pregnant women (@levallois2018public)(@jarvis2021lead), a critical gap remains in
understanding how lead concentrations in urban drinking water systems
fluctuate both spatially and temporally.

This paper pre-processed the raw data and the cleaned data was used for
visual representations and statistical analysis includes outliers
analysis, time series analysis and linear regression analysis.

The results from the analysis show that there is no clear seasonality,
which refers to recurring patterns or trends in data that repeat over
specific time intervals, and no significant autocorrelation, which
indicates a lack of correlation between current and past lead
concentrations. In other words, future lead concentrations do not depend
on previous lead levels, and there is insufficient evidence to suggest
that certain months or years consistently have higher trends of lead
concentrations.

However, some districts do have higher trend of lead concentration.
There are clusters of outliers within some specific districts and some
districts have higher mean lead concentrations than others. It is safe to
say that geographic locations and community water maintenance have an
effect on water safety.

This topic is significant because even low levels of lead exposure can
have serious health consequences, including developmental delays,
cognitive impairments and even detrimental effects(@c-1).

@talia is used for data analysis, @opendatatorotno for collected data
set, and @alexander2023telling for instructions and template.

The remainder of this paper is structured as follows. @sec-mydatasection
Data, provides a detailed explanation of the data set working with and the prepossessing done for further analysis. A discussion with the
measurement used when collecting the data is also included in @sec-measurement. Graphs and tables were used to emphasis the importance of the features in the data set.
@sec-mydiscussion Discussion, covers, @sec-health, to what extent will
Lead Concentration influence Health and how much concentration will be
categorized as “dangerous”. In @sec-district, the paper went into a detailed
analysis of how districts affect lead concentration. Graphs and outliers
analysis was used to present the result. Thirdly in @sec-time, the paper performs
Times Series analysis to analyze the lead concentration over time.
Lastly, @sec-future discussed weakness of the paper and some
further steps to further support the question.

# Data {#sec-mydatasection}

## Tools Used

R (R Core Team 2023) was used for the data analysis and the productions of all graphs and tables. The libraries: tidyverse (@o), dplyr (Hadley Wickham 2023), ggplot2 (@ggplot2), gt(@gt) , lubridtate(@lubridate), tidyr(@tidyr), and knitr (@knitr) were also used for organizing data and performing analysis.

## Data Set and Context
The data set "Non Regulated Lead Sample" by Open Data Toronto aims to
address the concerns Toronto residents have for their drinking water.
The data set contains information on lead concentrations in water
samples. Each row represents a water sample, with corresponding information
on the date, partial postal code where the sample was provided, and the
main variable of interest, lead concentration in parts per million
(ppm).

Lead exposure is an significant public health issue. High levels of lead
in drinking water can pose severe health risks, particularly to
vulnerable populations.(@w) Chronic exposure to lead can result in
developmental delays, neurological damage, and varous of other health
problems.(@w) Monitoring lead concentrations in water is crucial for
ensuring safe water quality and identifying areas requiring intervention
to effectively manage health risks.

There are similar data sets that address the problem of hard
water contamination. One example is from state of Michigan, USA, MIDHHS
collected similar sample with months, years and lead
concentrations(@m). Other data setanalyze the
different metals in water. However, this data set was chosen because lead is
a common metal in hard water with severe damage to public health
when overdose, and being a resident in city of Toronto, this is a
concern that needs to raise attention .

## Variables

The five variables in this paper: \_id, Sample Number, Sample Date,
PartialPostalCode and Lead Amount. While the first two are keys of this
data set for identification purpose, the remaining varaiables are of our interest to our analysis:

**Date**: Represents the date on which the water sample was collected,
can be used to detect trends and seasonal variations. It is evident from (@fig-showdate) that the data points are evenly distributed through time except during 2020, there is an empty period of time. This might be caused by COVID-19 quarantine, where the residence does not have the chance to volunteer their water sample.
```{r}
#| label: fig-showdate
#| fig-cap: "Graph of Dates Categorized by Year and Presented by Each Months"
#| echo: false
# Create a new data frame with extracted year and month
data_date <- data %>%
  transmute(
    Year = year(`Sample Date`),
    Month = month(`Sample Date`, label = TRUE, abbr = TRUE)
  )

# Plot with ggplot2
ggplot(data_date, aes(x = Year, y = Month)) +
  geom_jitter(width = 0.2, height = 0.2, color = blueishgrey, size = 3) + # Add jitter to separate points
  labs(title = "Distribution of Sample Dates by Year and Month",
       x = "Year",
       y = "Month") +
  theme_minimal()

```


**Partial Postal Code:** A truncated form of the postal code indicating
the general area where the water sample was taken, can be used to
identify potential geographical clusters. The (@fig-splitpostal) shows the partial postal code categorized by the first two characters and the y axis being the last character ranging from A-Z. It can be oberseved from (@fig-splitpostal) that the data points are not evenly distributed. This can be caused by sampling bias since it is a volunterring sampling but it can also be caused by the fact that some areas in City of Toronto are not present for residency.
```{r}
#| label: fig-splitpostal
#| fig-cap: "Graph of Partial Postal Code Splitting into First Two Characters and Last Character"
#| echo: false

# Create a new data frame with extracted components
data_postal <- data %>%
  transmute(
    FirstTwoChars = substr(district, 1, 2),
    LastChar = substr(district, 3, 3)
  )

# Plot the extracted components using ggplot2
ggplot(data_postal, aes(x = FirstTwoChars, y = LastChar)) +
  geom_jitter(width = 0.2, height = 0.2, color = blueishgrey, size = 3) + # Jitter points to spread them out for better visualization
  labs(title = "Relationship Between First Two and Last Characters of Postal Codes",
       x = "First Two Characters",
       y = "Last Character") +
  theme_minimal()

```

**Lead Concentration (ppm):** The measured amount of lead in the water
sample, expressed in parts per million(@fig-boxplot). This is the primary variable of
interest, as it directly relates to public health risks associated with
lead exposure.Additionally,     combinations of lead concentration with other variables can be furthur analyzed.
```{r}
#| label: fig-boxplot
#| fig-cap: "Boxplot showing the lead concentration above the saftey threshold"
#| echo: false

# Set the threshold value for the horizontal line
threshold_value <- 0.010011423

# Create the boxplot using ggplot2
ggplot(data, aes(y = `Lead Amount (ppm)`)) +
  geom_boxplot(fill = "skyblue") +
  geom_hline(yintercept = threshold_value, color = "red", linetype = "dashed") +
  labs(title = "Boxplot of Lead Amount (ppm) with Threshold Line",
       y = "Lead Amount (ppm)") +
  theme_minimal()

```

Other than the given variables, there are three additional variables
constructed during analysis: Months: Extracted from the Date variable to
analyze seasonal patterns in lead concentration. Years: Extracted from
the Date variable to evaluate trends over time and aggregate the data
for yearly analysis. District (First Two Digits of Postal Code): Created
from the Partial Postal Code to group data into broader geographic areas
within the city, aiding in the spatial analysis of lead concentration.

## Cleaning

For the purpose of data analysis the raw data set went through some
pre-processing: NA Values: All missing (NA) values were removed to
ensure accurate analysis. Lead Concentration: The Lead Amount (ppm)
variable was cleaned to remove any anomalies and ensure that values were
properly numeric. Date Variable Type: The Date variable was converted to
Date type in R to facilitate temporal analysis and aggregation by months
and years.

After cleaning the data looks like:

**Date(Date):** A Date type, ranges from 2014-01-01 to 2024-08-21,
covering over a decade of water sampling data.

**District(chr):** A Character type, represents various regions within
the city of Toronto based on the first two characters of the Partial
Postal Code.


**Lead Concentration(num):** A num type, which represents the summary
statistics of lead concentration typically include measures like the
minimum, maximum, mean, median, and standard deviation of lead levels
across the data set.

## Summary
A summary table(@tbl-summary) for num type variable lead amount.It is evident that the minimum and maximum value differs by 168,800 times, which is very significant. In (@gif-boxplot), it is evident that there are many outliers. Therefore, the assumption can be made that there are water samples that are very clean in terms of lead concentration and there are samples that are contaminated. It is clear that the mean and median differs a lot. Which means that the distribution of this data set is highly right skewed. The outliers will elevated mean lead concentration level in this data set.

```{r, echo=FALSE}
#| label: tbl-summary
#| tbl-cap: "Distribution Statistics Measurements of Lead Amount in ppm"
#| echo: false
# Example summary statistics
lead_summary <- summary(data$`Lead Amount (ppm)`)

# Convert summary to a data frame
summary_df <- data.frame(
  Metric = names(lead_summary),
  Value = as.numeric(lead_summary)
)

# Create a table using gt
gt_table <- summary_df %>%
  gt() %>%
  tab_header(
    title = "Summary of Lead Amount (ppm)"
  ) %>%
  cols_label(
    Metric = "Statistic",
    Value = "Value"
  ) %>%
  fmt_number(
    columns = "Value",
    decimals = 6
  )

# Print the gt table
gt_table

```
## Relationships
These relationships are upon further discussion in @sec-mydiscussion.

**Lead Concentration and Time (Months/Years)**: Analyzing lead levels
over time can reveal temporal trends, such as reductions in lead levels
following interventions, seasonal spikes, or gradual increases over
certain periods.

**Lead Concentration and Location (District)**: By examining lead
concentrations across districts, one can identify areas with
consistently high or low lead levels, indicating possible regional
differences in water quality or pipe infrastructure.

**Seasonality**: The monthly variable can help detect seasonal patterns,
such as increased lead concentration during warmer months due to factors
like water temperature affecting pipe corrosion.

**Temporal Trends**: By grouping data by year, any long-term trends,
such as a decrease in lead levels due to policy interventions, can be
detected.


## Measurement{#sec-measurement}
In this dataset, the aim is to explore in the concentration of lead in drinking water across different regions of a city and over time. The measurement process transforms a real-world phenomenon—lead contamination in water—into quantitative entries in a dataset that can be analyzed for trends. 

**Lead concentration:** Measured in parts per million (ppm), indicating the amount of lead in the water relative to its volume. This choice of unit is small enough to capture low but potentially harmful concentrations and is standard in water quality testing. This transformation from the chemical presence of lead to a numeric value is a key aspect of measurement.

**Sample Date:** The date the sample was taken is recorded to allow for analysis over time, detecting patterns.

**Partial Postal Code:** Postal code represents where the sample was collected, allowing comparisons between different geographic areas. This made it easy to get a result using data analysis to public health control organizations.

Some challenges are that this is a non-regulated program, therefore the measurement process is not controlled, decreasing accuracy and precision. However, this is minimized by providing a detailed instructions with the sample kit used to collect the water samples and the deviation will only affect Lead concentration. Date and district are accurate.

Choices about measurement such as the instruments used, units of measurement, and protocols for sample collection—affect how well the data set represents the real-world phenomenon of lead contamination.

# Discussion {#sec-mydiscussion}

## Lead Concentration influencing Health{#sec-health}

### WHO and Canadian federal guidance on lead concentration in water

The main variable examining in this paper is the Lead Amount in
ppm. The distribution of the data is highly right skewed with most of
its data being very small, as shown in (@fig-boxplot), most
data are almost 0 with outliers which are mostly less than 2 ppm
and 4 of them ranging from 3 to 8.44 which is the maximum value as shown
in the summary Table 1. Additionally observation can be verified
since the summary Table 1 shows that 75% of the data is below
0.001480. Government of Canada had specified that the maximum acceptable concentration (MAC) of drinking
water is 0.010mg/L(@c-1), which converts to 0.010011423
part/million(@c), this is shown by the red line in the
figure 1(@fig-boxplot).

### Percentage of data above this threshold

From calculations by R, there are 97.26% of the data collected indicating that
the lead amount in the water is safe. As shown in (@tbl-threshold), there are only 280 counts of water sample collected that is above MAC, which takes up only 2.74% of the whole data set. Even though this seems like a good result, those 280 measurements still effects people and their family's health since those lead concentrations will have negative effect on health(@w)(@c-1).
```{r, echo=FALSE}
#| label: tbl-threshold
#| tbl-cap: "Number and Percentage of data above Maximum Acceptable Concentration"
#| echo: false
# Define the threshold
threshold <- 0.010

# Subset of samples above the threshold
high_lead_samples <- data[data$`Lead Amount (ppm)` > threshold, ]

# Calculate the count and percentage
count_above_threshold <- nrow(high_lead_samples)
percentage_above_threshold <- (count_above_threshold / nrow(data)) * 100

# Create a summary table
summary_table <- data.frame(
  Metric = c("Count of Samples Above Threshold", "Percentage of Samples Above Threshold"),
  Value = c(count_above_threshold, round(percentage_above_threshold, 2))
)

# Use gt to create a formatted table
summary_table %>%
  gt() %>%
  tab_header(
    title = "Summary of Lead Concentration Above Threshold"
  ) %>%
  fmt_number(
    columns = "Value",
    decimals = 2
  )


```

### plot of the data above this threshold
To visualize those data above MAC, data below MAC and above 2 were filtered out, because from (@fig-boxplot), it is clear that there are only 4 data points above 2, therefore, with that in mind it is safe to temporarily ignore to get a scale that will better show the most data points above MAC. A scatterplot was used with x axis being date and y axis being lead concentration give a graph that spreads out lead concentrations instead of on a 1-dimensional line. From (@fig-between) this indicates that there are clusters of data points below 0.25 and for higher lead concentrations, there are less data points and more spread out than those below 0.25.

```{r}
#| label: fig-between
#| fig-cap: "Data Between 2ppm and Maximum Acceptable Concentration Representing By Date"
#| echo: false
# Define the threshold value
threshold <- 0.010011423
upper_threshold = 2

# Filter data for values above the threshold
data_above_threshold <- data %>%
  filter(upper_threshold >`Lead Amount (ppm)` & `Lead Amount (ppm)` > threshold)

# Plot the data using ggplot2 (scatter plot example)
ggplot(data_above_threshold, aes(x = `Sample Date`, y = `Lead Amount (ppm)`)) +
  geom_point(color = "red") +  # Use red points to highlight threshold exceedance
  labs(title = "Lead Levels Above Threshold",
       x = "Sample Date",
       y = "Lead Amount (ppm)") +
  theme_minimal()

```

## How districts affect Lead Concentration{#sec-district}

### graphs of postal code vs lead concentrations

Nine boxplot (@fig-boxplotdistrict) each representing a district, which is the constructed variable from the first two letters of partial postal code. Again, most data points are around 0, and the outliers streched the scale so that this way we can easily analyze the outliers. The districts containing outliers were also shown in (@fig-boxplotdistrict). It is clear that the "M4" has two data points above 8 ppm, as
well as a cluster between 0 and 1 ppm. "M6" also have many data between 0 and
2ppm. It is safe to assume that these two districts have water sources with high lead concentrations.
```{r}
#| label: fig-boxplotdistrict
#| fig-cap: "Boxplot Showing Outliers Categorized by District"
#| echo: false


# Create a new data frame with District extracted
data_district <- data.frame(
  District = substr(data$PartialPostalCode, 1, 2),
 lead = data$`Lead Amount (ppm)`
)

# Create the boxplot using ggplot2 without modifying the original data
ggplot(data_district, aes(x = District, y = lead)) +
  geom_boxplot(fill = "black", color = blueishgrey, outlier.color = "red", outlier.shape = 16) +
  labs(title = "Lead Levels by District",
       x = "District (First Two Characters of Postal Code)",
       y = "Lead Amount (ppm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  # Center and bold title
    axis.title.x = element_text(size = 14, face = "bold"),  # Bold x-axis title
    axis.title.y = element_text(size = 14, face = "bold"),  # Bold y-axis title
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12)
  )


```

### extract outliers above the threshold

Verification will be conducted which district have a higher overall lead concentration by calculating the mean lead concentration when still categorize the partial postal codes by only the
first two numbers. Since outliers might influence the mean, especially
when in our dataset we are dealing with median 0.000473 and outliers as
high as 8ppm, the outliers above 2 ppm were omitted before caculating the mean lead concentration.

### mean value of lead concentration for different districts

The bargraph (@fig-meandistrict) compares the mean values of lead
concentration without outliers above 2 ppm and categorized by only the
first two characters of the partial postal code. It is evident that the top
two district of the highest lead concentration in water is M9 and M6 respectively. This shows that without the data above 2 ppm, M4 does not have a overall high lead concentration, but only some very high outliers. However, M9 and M6 will be the two most dangerous district in Toronto in terms of lead concentration in water. Even though all the mean values shown in (@tbl-meandistrict) are all below MAC by Canadian Government, the M9 and M6 district might need more attention in the topic of water saftey given that the rest districts perform better especially M8.

```{r}
#| label: tbl-meandistrict
#| tbl-cap: "Specific Numbers of the Mean Lead Concentration for each District"
#| echo: false


# Remove rows with LeadAmountPPM > 4
filtered_data <- subset(data, `Lead Amount (ppm)` <= 2)

# Categorize by the first two characters of PartialPostalCode
filtered_data$District <- substr(filtered_data$PartialPostalCode, 1, 2)

# Step 3: Calculate the mean of LeadAmountPPM by District
mean_lead_by_district <- aggregate(`Lead Amount (ppm)` ~ District, data = filtered_data, FUN = mean)


# Create the table using gt
mean_lead_by_district %>%
  gt() %>%
  tab_header(
    title = "Mean Lead Concentration by District"
  ) %>%
  cols_label(
    District = "District",
    `Lead Amount (ppm)` = "Mean Lead Amount (ppm)"
  ) %>%
  fmt_number(
    columns = `Lead Amount (ppm)`,
    decimals = 5
  )

```
\newpage
```{r}
#| label: fig-meandistrict
#| fig-cap: "Mean Lead Concentration Categorized by First Two Characters from Partial Postal Code"
#| echo: false
# Step 4: Construct a bar graph using ggplot2
ggplot(mean_lead_by_district, aes(x = District, y = `Lead Amount (ppm)`)) +
  geom_bar(stat = "identity", fill = blueishgrey) +
  labs(title = "Mean Lead Concentration by District",
       x = "District (First Two Characters of Postal Code)",
       y = "Mean Lead Amount (ppm)") +
  theme_minimal()
```


## How months and year affect Lead Concentration{#sec-time}

### Times Series analysis for autocorrelation{#sec-timeseries}

In this dataset, the variables are time and lead amount,
therefore,times series analysis can be performed. As shown in
(@fig-timeseries), it is evident that except for some sharp increase
in 2014, 2019, 2021, 2023, the overall trend is stays around 0 and
remains positive. There are no increasing or decreasing trend overtime and no obvious seasonality present.

From the ACF and PACF figures, it can be concluded that the previous values
of lead concentrations does not impact future values. For ACF, when lag
is 0, ACF is always 1 but for all other lags in ACF(@fig-ACF) and
PACF(@fig-PACF) they are all within the dotted blue lines which is
the evidence that there are no significant correlation at any lags present in the data.

This observation is reasonable because the samples are collected in
different district in Toronto, and it is safe to assume that one
incident of a very high lead concentration might again only be outliers
instead of due some factors that will effect the overall water quality in
Toronto. 

Lead Concentration in water can vary overtime(@w), therefore the high lead concentration outliers might not be caused by correlation instead, "a probability-based adaptive sampling plan should be used to access exposure(@w)".

```{r}
#| label: fig-timeseries
#| fig-cap: "Lead Levels Changed Over Time for Times Series Analysis"
#| echo: false

# Create a line plot using ggplot2
ggplot(data, aes(x = `Sample Date`, y = `Lead Amount (ppm)`)) +
  geom_line(color = blueishgrey, linewidth = 0.5) +
  labs(title = "Lead Levels Over Time",
       x = "Date",
       y = "Lead Amount (ppm)") +
  theme_minimal()


```

```{r}
#| label: fig-ACF
#| fig-cap: "Autocorrelation Function for Lead Concentrations"
#| echo: false
# Autocorrelation function (ACF)
# Calculate ACF (autocorrelation function)
acf_result <- acf(data$`Lead Amount (ppm)`, na.action = na.omit, plot = FALSE)

# Extract the lags and acf values from the result
acf_data <- data.frame(
  Lag = acf_result$lag,
  ACF = acf_result$acf
)

# Create the ACF plot using ggplot2
ggplot(acf_data, aes(x = Lag, y = ACF)) +
  geom_bar(stat = "identity", fill = "white", color = "black") +
  geom_hline(yintercept = c(0.2, -0.2), linetype = "dashed", color = "blue") +  # Add reference lines
  labs(title = "Autocorrelation of Lead Levels",
       x = "Lag",
       y = "Autocorrelation") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  # Center and bold title
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

```

```{r}
#| label: fig-PACF
#| fig-cap: "Partial Autocorrelation Function for Lead Concentrations"
#| echo: false
# Partial Autocorrelation function (PACF)
pacf = pacf(data$`Lead Amount (ppm)`, na.action = na.omit, main="Partial Autocorrelation of Lead Levels")


```

### graphs of months vs mean lead concentration
With all data points, there are no obvious trends over time. The examination is made closely for the constructed variable month from Date. (@fig-month) is the boxplot showing the outliers, but there are not enough evidence to show that there are specific month that will have a higher lead concentration. Therefore, a similar filtering is applied to data larger than 2 ppm.
```{r}
#| label: fig-month
#| fig-cap: "Boxplots that Shows the Outliers Categorized by Months"
#| echo: false
# Extract month as a factor for visualization
data_by_month <- data %>%
  mutate(Month = month(`Sample Date`, label = TRUE, abbr = TRUE))

# Create a boxplot of lead concentration by month with red outliers
ggplot(data_by_month, aes(x = Month, y = `Lead Amount (ppm)`)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, fill = blueishgrey) +
  labs(title = "Lead Concentration by Month",
       x = "Month",
       y = "Lead Amount (ppm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )

```

Bar graph (@fig-meanmonth) shows the mean lead concentration for each month in the year. April and September has higher lead concentrations that is about 500% of the other months. However, this graph does not provide enough evidence to prove seasonality. It is necessary to examine whether April and September for every year will repeatedly have higher measured lead concentration. This will be further discussed in @sec-future.


```{r}
#| label: fig-meanmonth
#| fig-cap: "Mean Lead Concentration Without Data Greater than 2 ppm Categorized by Months"
#| echo: false
# Extract month and calculate mean lead concentration by month
mean_lead_by_month <- data %>%
  mutate(Month = month(`Sample Date`, label = TRUE, abbr = TRUE)) %>%
  group_by(Month) %>%
  summarise(mean_lead = mean(`Lead Amount (ppm)`))

# Create the plot of mean lead concentration by month
ggplot(mean_lead_by_month, aes(x = Month, y = mean_lead)) +
  geom_bar(stat = "identity", fill = blueishgrey) +
  labs(title = "Mean Lead Concentration by Month",
       x = "Month",
       y = "Mean Lead Amount (ppm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )

```

### graphs of years vs mean lead concentration
Similar to months,variable called year was also constructed. However, both the boxplots and the mean values without data below 2 ppm does not show trend from year to year. This aligns with our conclusion in @sec-timeseries.
```{r}
#| label: fig-year
#| fig-cap: "Boxplots that Shows the Outliers Categorized by Year"
#| echo: false
# Create a new data frame with the extracted year
data_by_year <- data.frame(
  Year = year(data$`Sample Date`),
  lead = data$`Lead Amount (ppm)`
)

# Create a ggplot with lead concentration categorized by years
ggplot(data_by_year, aes(x = factor(Year), y = lead)) +
  geom_boxplot(fill = "skyblue", color = blueishgrey,outlier.color = "red") +
  labs(title = "Lead Concentration by Year",
       x = "Year",
       y = "Lead Amount (ppm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )

```
```{r}
#| label: fig-meanyear
#| fig-cap: "Mean Lead Concentration Without Data Greater than 2 ppm categorized by Year"
#| echo: false
# Filter data with threshold of 2 ppm
filtered_data <- data %>%
  filter(`Lead Amount (ppm)` <= 2)

# Extract year and calculate mean lead concentration by year
mean_lead_by_year <- filtered_data %>%
  mutate(Year = year(`Sample Date`)) %>%
  group_by(Year) %>%
  summarise(mean_lead = mean(`Lead Amount (ppm)`))

# Create the plot of mean lead concentration by year
ggplot(mean_lead_by_year, aes(x = factor(Year), y = mean_lead)) +
  geom_bar(stat = "identity", fill = blueishgrey, color = blueishgrey) +
  labs(title = "Mean Lead Concentration by Year",
       x = "Year",
       y = "Mean Lead Amount (ppm)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )

```

## Weaknesses and next steps{#sec-future}

A key weakness of the current dataset is that it originates from a variety of water sources and is not necessarily limited to direct drinking water. Since the water sample collection process was not controlled, it is uncertain whether all samples reflect tap water quality. Despite this, the guidelines used for analysis are intended for drinking water safety, which introduces a potential mismatch between the data and its intended use. The program is non-regulated, leaving the sampling process up to residents and without any standardization; as a result, the reliability and comparability of the data across different samples may be compromised.

Considering these weaknesses, it is important to recognize the role of pre-treatment, such as water filtration systems, in affecting lead concentration. This necessitates the documentation of whether samples are taken before or after filtration, which could significantly influence the interpretation of lead levels in relation to health guidelines. To improve the reliability of future data collection, a probability-based adaptive sampling plan should be considered. This approach would ensure more representative sampling of tap water from various districts, providing a better understanding of the exposure to lead in drinking water. Additionally, further analysis on the seasonality of lead levels could reveal patterns influenced by environmental factors such as temperature, water flow rates, and seasonal water treatment processes. Identifying these trends is crucial for developing targeted interventions to reduce lead exposure throughout the year.

\newpage

\appendix

# Appendix {.unnumbered}
The color used in graphs :
blueishgrey <- rgb(96, 139, 182, maxColorValue = 255) is the color of lead in reality.

# Additional data details

\newpage

# References
